---
title: 'Project 2: Modelling daily cycling demand in Edinburgh'
author: 'Group members: [Karthik Sutheeshkumar], [Kiet Lam], [Leon]'
date: 'Word count: [0 words]'
output:
  pdf_document:
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
  html_document:
    df_print: paged
header-includes:
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{amsmath}
- \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,    # This hides all code from the main body 
  message   = FALSE, 
  warning   = FALSE
)
source("Proj2code.R") # Runs all models and creates tables/plots in the background
```

# Introduction


<!--
Required content (remove before submission):
3-5 sentences covering:
  - What the dataset is and where it comes from
  - What the modelling goal is
  - Why this is useful for transport planning in Edinburgh
Do NOT describe the report structure here.
-->

Edinburgh operates a network of automatic cycling counters at strategic locations across the city that records the total number of cyclists passing each sit daily, The resulting dataset contains approximately 2,180 observations spanning 2020-2025,covering a period that includes the COVID-19 pandemic and a gradual return to normal travel patterns. Using this data, a sequence of linear regression models of increasing  complexity, evaluate them rigorously to explain how daily cycling demand varies with weather conditions, day of week, seasonality, and longer-term trend, evaluating each model using leave-one-year-out cross validation. The analysis is guided by three questions of practical interest to Edinburgh City Council: how well the model predicts demand across different seasons, how sensitive cycling is to temperature, and what the long-term trend implies for infrastructure 
planning over the next decade.


# Exploratory data analysis

<!--
Required content:
  1. Summary statistics table for count, temp_mean, temp_min, temp_max.
  2. Time series of daily counts 2020-2025 with a smoother.
  3. Boxplots of count by month and by day of week.
  4. Scatter plot of count vs temp_mean.
  5. At least one further plot of your choice highlighting a notable feature.
Every figure must have a caption and be explicitly discussed in the text.
-->

**Summary statistics table**

```{r,echo=FALSE}
kable_summary_stats
```

The above table summarises the key numeric variables across the 2100 daily observations. Daily cyclist counts average around 10,816 but with a large standard deviation of 4729 reflecting strong seasonal variation- the minimum of 1,264 and maximum of 43,864 confirm that demand varies dramatically across the year. Edinburgh's climate is characterised by mild but cool conditions, with mean daily temperatures averaging 9.7°C and ranging from -8.1°C to 23°C, typical for a Scottish city. The wide spread in temperature variables suggests that weather is likely to be an important predictor of cycling demand, which motivates its inclusion in all models. Notably, the minimum temperature recorded (-14.0°C) is well below freezing, conditions under which cycling activity would be expected to drop sharply.


**Time series of daily counts with a smoother **

```{r,echo=FALSE}
eda_time_series_plot
```

The above figure shows the daily cyclist counts across the full observation period from 2020-2025, with a smoother overlaid to highlight the trend, filtering out day-to-day variability. We can see the obvious pattern that the count climbs every summer and fall back each winter, and this repeats every year.There is a visible dip around 2020-2021, which is unsurprising given the COVID-19 lockdowns during that period. After that, counts reach their highest levels around 2022-23. The drop at the end of the observation period coincides with the data 
ending in early 2026, making it difficult to distinguish a seasonal effect from a genuine trend change. Certain days stands out in count, especially one around 2022 that exceeds 40,000 cyclists a day, which probably reflects an unusually warm day.


**Boxplots of count by month and by day of week **

```{r}
eda_month_boxplot
```

The boxplot by month confirms the seasonal pattern seen in the time series. Counts 
are lowest in January and December, with medians around 7,000-8,000 cyclists per day, rising steadily through spring and peaking between May and August. Winter months see roughly half the cycling activity of summer months, which has clear implications for seasonal infrastructure planning.


```{r,echo=FALSE}
eda_dow_boxplot
```

Cycling demand is noticeably higher on weekdays than weekends, with Tuesday to Friday showing the highest medians of around 11,000-12,000 cyclists per day. Sunday and Monday stand out as the lowest, with medians closer to 8,000-9,000, suggesting that commuting is a major driver of cycling demand in Edinburgh. 

**Scatter plot of count vs temp_mean**

```{r,echo=FALSE}
eda_temp_scatter
```

The above figure shows a clear positive but non-linear relationship between mean temperature and daily cyclist counts. Counts rise steeply at lower temperatures before leveling off above around 10–12°C, motivating the quadratic temperature term used in M2 and M3. The spread of counts also increases with temperature, suggesting heteroscedasticity and supporting the log-transform applied in M3.

**Showing relation between count and mean temperature during COVID and post COVID period**


```{r,echo=FALSE}
eda_covid_temp_plot
```

The above graph reveals an interesting and somewhat counterintuitive pattern — cycling counts during the COVID period (2020-2021) were actually higher at warmer 
temperatures compared to post-COVID years, with the red smoother sitting well 
above the blue at temperatures above 10°C. This likely reflects that during 
lockdowns, cycling shifted from commuting to recreational use, with people more 
willing to cycle on pleasant days but much less likely to cycle in cold or wet 
conditions, explaining the steeper temperature sensitivity. At cold temperatures 
the pattern reverses, with post-COVID counts higher, consistent with commuters 
cycling regardless of weather. This difference in behaviour between the two periods suggests that a COVID indicator variable may improve model fit, and motivates its 
inclusion as a candidate extension in M3.


# Baseline model M0


$$

\begin{equation}
\text{count}_i = \beta_0 
+ \beta_1 \cdot \text{temp_mean}_i 
+ \beta_2 \cdot \text{weekend}_i 
+ \sum_{m=2}^{12} \beta_m \cdot \mathbf{1}(\text{month}_i = m) 
+ \varepsilon_i, 
\quad \varepsilon_i \sim N(0, \sigma^2)
\end{equation}

$$

```{r,echo=FALSE}
kable_m0_coeff_table
```

Table shows the estimated coefficients for M0. All four predictors are highly 
significant (p < 0.001). Each 1°C increase in mean temperature is associated with 
approximately 247 additional cyclists per day, while weekend days see around 1,247 
fewer cyclists than weekdays, consistent with the commuter-driven pattern observed 
in the EDA. The negative month coefficient (-247) implies that counts decline on 
average as the month number increases, though this is a crude approximation since 
the relationship between month and demand is not linear.


```{r,echo=FALSE}
m0_resid_fitted_plot + m0_qq_plot
```

The residuals versus fitted plot shows that the spread of residuals grows 
noticeably with fitted values, indicating heteroscedasticity — a violation 
of the constant variance assumption. The QQ plot reveals a heavy right tail, with standardised residuals curving sharply above the reference line at high quantiles, reflecting the right-skewed distribution typical of non-negative count data. Both issues suggest that M0's Gaussian assumption is inadequate. 

M0 captures the broad direction of key predictors — temperature, weekend effect, 
and seasonality but treats month as numeric, forcing a linear relationship across months that does not reflect the true seasonal pattern seen in the EDA. It also 
omits day of week entirely and includes no long-term trend. These limitations 
motivate M1, which switches month to a factor, adds day of week as a factor, and 
introduces a linear trend term to capture drift in cycling demand over 2020–2025.



<!--
Required content:
  1. State M0 mathematically using a numbered equation.
  2. Coefficient table (use kable, not raw output).
  3. Residuals vs fitted and normal QQ diagnostic plots, side by side.
  4. Brief assessment: what does M0 get right and wrong? 
     What motivates M1?
-->

# Proposed models M1--M3

<!--
For each model:
  - State the model mathematically.
  - Justify every new term BEFORE showing results.
  - Show diagnostic plots and comment on improvement over the previous model.
-->

## Model M1: Trend and factor 

$$
\text{count}_i = \beta_0 
+ \beta_1 \cdot \text{temp_mean}_i 
+ \beta_2 \cdot \text{weekend}_i 
+ \beta_3 \cdot \text{trend}_i 
+ \sum_{m=2}^{12} \gamma_m \cdot \mathbf{1}(\text{month}_i = m) 
+ \sum_{d=2}^{7} \delta_d \cdot \mathbf{1}(\text{dow}_i = d) 
+ \varepsilon_i, \quad \varepsilon_i \sim N(0, \sigma^2)
$$


M1 extends the baseline model in three ways, each motivated by clear limitations 
identified in M0. 

First, month is switched from a numeric predictor to a factor, giving each month 
its own intercept. The boxplot in Figure 2 shows that the relationship between 
month and demand is not monotonic — counts peak in summer and fall in winter but 
do not follow a straight line across month numbers 1 to 12, so forcing a single 
slope as M0 does is inappropriate.

Second, day of week is added as a factor. The boxplot in Figure 3 showed a clear 
difference between weekdays and weekends, but the weekend binary in M0 groups 
Saturday and Sunday together. Since Saturday behaves differently from Sunday — 
likely reflecting a mix of leisure and residual commuting trips — a full factor 
allows each day to have its own effect.

Third, a linear trend term is added, defined as the number of days since 1 January 
2020. The time series in Figure 1 suggests that underlying demand has shifted over 
the 2020–2025 period, and without a trend term the model has no way to capture 
this drift, potentially biasing all other coefficient estimates

Diagnostics of m1

```{r,echo=FALSE}
m1_resid_fitted_plot + m1_qq_plot
```

Compared to M0, M1 shows some improvement — the fitted values span a wider range, 
suggesting the additional predictors (trend, factor month, factor day of week) give the model more flexibility to capture variation in demand. However, the core issues persist. The residuals versus fitted plot still shows increasing spread at higher 
fitted values, indicating heteroscedasticity remains unresolved. The QQ plot 
continues to show a pronounced right tail, with standardized residuals reaching 
above 6 at the upper end, reflecting the right-skewed nature of the count data 
that a Gaussian model struggles to accommodate. These issues motivate further 
extensions in M2 and M3.


## Model M2: Nonlinear temperature


$$
\text{count}_i = \beta_0 
+ \beta_1 \cdot \text{temp_mean}_i
+ \beta_2 \cdot \text{temp_mean}^2_i
+ \beta_3 \cdot \text{weekend}_i 
+ \beta_4 \cdot \text{trend}_i 
+ \sum_{m=2}^{12} \gamma_m \cdot \mathbf{1}(\text{month}_i = m) 
+ \sum_{d=2}^{7} \delta_d \cdot \mathbf{1}(\text{dow}_i = d) 
+ \varepsilon_i, \quad \varepsilon_i \sim N(0, \sigma^2)
$$



M2 extends M1 by replacing the linear temperature term with a quadratic, 
adding both temp_mean and temp_mean² as predictors. The scatter plot in 
Figure 3 showed that the relationship between temperature and cycling demand 
flattens out above around 10°C rather than increasing at a constant rate, 
which a linear term cannot capture. Including the quadratic term allows the 
model to reflect this diminishing sensitivity to temperature at higher values, 
which is also physically plausible — once conditions are warm enough, further 
increases in temperature have a smaller effect on whether people choose to cycle.


```{r,echo=FALSE}
m2_resid_fitted_plot+m2_qq_plot
```

M2 shows marginal improvement over M1 — the residuals versus fitted plot appears 
slightly more evenly spread, suggesting the quadratic temperature term has captured some additional structure in the data. The lower tail of the QQ plot also follows 
the reference line more closely than in M1. However, the right tail remains heavily skewed, with standardised residuals exceeding 7 at the upper end, worse than M1. 
Heteroscedasticity also persists, with residual spread still growing at higher 
fitted values. The consistently heavy right tail across M0, M1, and M2 strongly 
suggests that a Gaussian model is fundamentally inappropriate for these count data, and motivates the log transformation of the response in M3.



## Model M3: [Log-Transformed Response]

$$
\log(\text{count}_i + 1) = \beta_0 
+ \beta_1 \cdot \text{temp\_mean}_i 
+ \beta_2 \cdot \text{temp\_mean}_i^2 
+ \beta_3 \cdot \text{weekend}_i 
+ \beta_4 \cdot \text{trend}_i 
+ \sum_{m=2}^{12} \gamma_m \cdot \mathbf{1}(\text{month}_i = m) 
+ \sum_{d=2}^{7} \delta_d \cdot \mathbf{1}(\text{dow}_i = d) 
+ \varepsilon_i, \quad \varepsilon_i \sim N(0, \sigma^2)
$$

M3 keeps all predictors from M2 but log-transforms the response variable, 
fitting log(count + 1) rather than count directly. The diagnostic plots for 
M0, M1, and M2 consistently showed two problems: a heavy right tail in the 
QQ plot and residual spread increasing with fitted values. Both are classic 
symptoms of right-skewed, non-negative count data being modelled with a 
Gaussian assumption. The log transformation directly addresses this — by 
compressing large values, it reduces skewness in the response and stabilises 
variance, making the Gaussian assumption more plausible. The +1 is added 
before taking the log to handle any zero counts, since log(0) is undefined.


```{r,echo=FALSE}
m3_resid_fitted_plot + m3_qq_plot
```

The diagnostic plots for M3 show a marked improvement over all previous models. 
The residuals versus fitted plot no longer shows the growing spread seen in M0 
through M2 — residuals are now scattered fairly evenly around zero across the 
full range of fitted values, suggesting the log transformation has largely 
resolved the heteroscedasticity. The QQ plot is dramatically better, with 
standardised residuals following the reference line closely across almost the 
entire range, compared to the severe right tail deviation seen in M2 where 
residuals exceeded 7. The tails still deviate slightly at both ends, indicating 
some remaining non-normality, but this is a substantial improvement. Overall, 
the log transformation successfully addressed the two core issues identified in 
M0 through M2 — right skew and non-constant variance — making the Gaussian 
assumption considerably more plausible for M3 than for any of the earlier models.



# Evaluation
## Statistical Performance

<!--
Required content:
  1. Table 1: LOYO CV scores for M0-M3.
  2. Table 2: CV RMSE and DS by month for your final model.
  3. Discussion: do the scores agree on which model is best? 
     If not, explain why and state your final model choice explicitly.
-->

## Practical Performance

<!--
One subsection per planning question (5.1-5.5 from the assignment brief).
Each must include mandatory fixed-format numbers and real-world interpretation.
-->

# Discussion and limitations

<!--
3-5 paragraphs covering:
  - Whether model assumptions appear satisfied.
  - Key limitations of the model and data.
  - What ideas would improve prediction.
  - Whether findings generalise to other cities or future years.
Do not introduce new results here.
-->

# Appendix: Code

All analysis code is in `code.R` and is reproduced below.

```{r print-code, echo=TRUE, eval=FALSE, code=readLines("code.R")}
```


```{r}
summary_stats
```

