---
title: 'Project 2: Modelling daily cycling demand in Edinburgh'
author: 'Group members: [Karthik Sutheeshkumar], [Kiet Lam], [Leon]'
date: 'Word count: [0 words]'
output:
  bookdown::pdf_document2:
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: true
    toc: false
  html_document:
    df_print: paged
header-includes:
- \usepackage{booktabs}
- \usepackage{float}
- \usepackage{amsmath}
- \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,    # This hides all code from the main body 
  message   = FALSE, 
  warning   = FALSE 
)
source("Proj2code.R") # Runs all models and creates tables/plots in the background
```

# Introduction

 
<!--
Required content (remove before submission):
3-5 sentences covering:
  - What the dataset is and where it comes from
  - What the modelling goal is
  - Why this is useful for transport planning in Edinburgh
Do NOT describe the report structure here.
-->

Edinburgh operates a network of automatic cycling counters at strategic locations across the city that records the total number of cyclists passing each site daily, The resulting dataset contains 2,180 observations spanning from 2020 to 2025, covering a period that includes the COVID-19 pandemic and a gradual return to normal travel patterns. Using this data, we evaluate a sequence of linear regression models of increasing complexity rigorously to explain how daily cycling demand varies with weather conditions, day of week, seasonality, and longer-term trend, evaluating each model using leave-one-year-out cross validation. The analysis is guided by three questions of practical interest to Edinburgh City Council: how well the model predicts demand across different seasons, how sensitive cycling is to temperature, and what the long-term trend implies for infrastructure planning over the next decade.
  

# Exploratory data analysis

Three derived variables are created before modelling: month as an ordered factor 
(levels 1–12), day of week with explicit levels Sunday to Saturday, and trend as 
integer days since 1 January 2020 to capture linear drift in cycling demand.



<!--
Required content:
  1. Summary statistics table for count, temp_mean, temp_min, temp_max.
  2. Time series of daily counts 2020-2025 with a smoother.
  3. Boxplots of count by month and by day of week.
  4. Scatter plot of count vs temp_mean.
  5. At least one further plot of your choice highlighting a notable feature.
Every figure must have a caption and be explicitly discussed in the text.
-->

```{r,echo=FALSE}
kable(summary_stats,
      caption = "Summary statistics for daily cyclist count and temperature variables.",
      booktabs = TRUE,format = "latex",
      linesep = "") %>%
  kable_styling(latex_options = "hold_position")
```

Table 1 summarises the key numeric variables. Daily counts average 10,816 with a 
standard deviation of 4,729, confirming strong seasonal variation — the range of 
1,264 to 43,864 reflects dramatically different demand across the year. Mean 
temperature averages 9.7°C and ranges from -8.1°C to 23°C, typical for Edinburgh, 
with the full temperature spread of -14°C to 31°C suggesting weather will be an 
important predictor of cycling demand.

```{r fig-ts, echo=FALSE, fig.cap=" Time series of daily cyclist counts in Edinburgh (2020–2025) with a smoother.",fig.width=7, fig.height=4,}
eda_time_series_plot
```

Figure \@ref(fig:fig-ts) shows daily cyclist counts from 2020 to 2025 with a smoother overlaid. 
The seasonal pattern is clear — counts climb each summer and fall each winter 
consistently across all years. A dip around 2020-2021 reflects COVID-19 lockdowns, 
after which counts peak around 2022-23. One day around 2022 exceeds 40,000 cyclists, 
likely an unusually warm day. The drop at the end coincides with the data ending in 
early 2026, making it difficult to distinguish a seasonal effect from a genuine trend 
change. 

## Boxplots of count by month and by day of week 

```{r fig-boxplotmonth,echo=FALSE, fig.cap="Distribution of daily cyclist counts by month.",fig.width=6, fig.height=2.25}
eda_month_boxplot
```

Figure \@ref(fig:fig-boxplotmonth) confirms the seasonal pattern seen in the time series. Counts are lowest in January and December, with medians around 10,500-11,500 cyclists per day, rising steadily through spring and peaking between May and August. Winter months see roughly half the cycling activity of summer months, which has clear implications for seasonal infrastructure planning.


```{r fig-boxplotdow,echo=FALSE, fig.cap="Distribution of daily cyclist counts by day of week",fig.width=6, fig.height=3}
eda_dow_boxplot
```

Figure \@ref(fig:fig-boxplotdow) suggests that cycling demand is higher on weekdays than weekends, with Tuesday to Friday showing the highest medians of around 11,000-12,000 cyclists per day. Sunday and Monday stand out as the lowest, with medians closer to 8,000-9,000, suggesting that commuting is a major driver of cycling demand in Edinburgh. 

## Scatter plot of count vs temp_mean 

```{r fig-scattertemp,echo=FALSE, fig.cap="Daily cyclist count versus mean temperature",fig.width=6, fig.height=3}
eda_temp_scatter
```

Figure \@ref(fig:fig-scattertemp) shows a clear positive but non-linear relationship between mean temperature and daily cyclist counts. Counts rise steeply at lower temperatures before leveling off above around 10–12°C, suggesting a quadratic relationship for temperature term. The spread of counts also increases with temperature, suggesting heteroscedasticity.

## Showing relation between count and mean temperature during COVID and post COVID period


```{r fig-covidscatter,echo=FALSE, fig.cap="Daily cyclist count versus mean temperature by COVID and post-COVID period (2020–2025)",fig.width=6, fig.height=3}
eda_covid_temp_plot
```

 Figure \@ref(fig:fig-covidscatter) reveals a counterintuitive pattern — during COVID (2020-2021), counts 
were higher at warmer temperatures than post-COVID, likely because lockdowns 
shifted cycling from commuting to recreational use, making demand more 
temperature-sensitive. At cold temperatures the pattern reverses, with post-COVID 
counts higher, consistent with commuters cycling regardless of weather. This 
motivates including a COVID indicator as a candidate extension in M3.


# Baseline model M0

In all of the models that we will fit, the response variable is $y_i = \text{count}_i$, the total 
daily cyclist count on date $i$, for each day $i$ from 1 January 2020 to 31 December 
2025. Four models of increasing complexity are considered: M0 (baseline with numeric 
month), M1 (adds linear trend and factor covariates), M2 (replaces linear temperature 
with a quadratic term), and M3 (log-transforms the response to address non-normality). 
The common predictors across all models are defined as follows:

\begin{align*}
x_{1i} &= \text{temp\_mean}_i \quad \text{(daily mean air temperature in °C)} \\
x_{2i} &= \text{weekend}_i \quad \text{(binary indicator: 1 if Saturday or Sunday, 0 otherwise)} \\
m_i &\in \{1, \ldots, 12\} \quad \text{(month: numeric in M0, factor in M1--M3)} \\
t_i &= \text{trend}_i \quad \text{(integer days since 2020-01-01, where $t_i = 0$ on 1 January 2020)} \\
d_i &= \text{dow}_i \quad \text{(day of week factor, reference level: Sunday)}
\end{align*}
All models assume $\varepsilon_i \sim N(0, \sigma^2)$.

The baseline model M0, fitted for each day $i$ from 1 January 2020 to 
31 December 2025, is specified as:

$$
y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \beta_3 m_i + \varepsilon_i, 
\quad \varepsilon_i \sim N(0, \sigma^2)
$$


```{r,echo=FALSE}
kable(m0_coef_table,
      caption = "Coefficients table for baseline model M0.", booktabs = TRUE,format ="latex",
      linesep = "")%>%
  kable_styling(latex_options = "hold_position")
```

Table 2 shows the estimated coefficients for M0. All four predictors are highly significant (p < 0.001). Each 1°C increase in 
temperature is associated with approximately 247 additional cyclists per day, 
weekend days see around 1,247 fewer cyclists than weekdays, and the negative 
month coefficient (-247) implies demand declines as month number increases — 
a crude approximation since the true seasonal pattern is nonlinear.


```{r fig-m0diagnostics,echo=FALSE,fig.cap="Diagnostics for M0",fig.width=7, fig.height=3}
m0_resid_fitted_plot + m0_qq_plot
```

The residuals versus fitted plot shows residual spread growing with fitted values, 
indicating heteroscedasticity. The QQ plot reveals a heavy right tail with 
standardized residuals curving sharply above the reference line, suggesting M0's 
Gaussian assumption is inadequate.

M0 captures the broad direction of key predictors but treats month as numeric, 
omits day of week entirely, and includes no long-term trend. These limitations 
motivate M1, which switches month to a factor, adds day of week as a factor, and 
introduces a linear trend term.


<!--
Required content:
  1. State M0 mathematically using a numbered equation.
  2. Coefficient table (use kable, not raw output).
  3. Residuals vs fitted and normal QQ diagnostic plots, side by side.
  4. Brief assessment: what does M0 get right and wrong? 
     What motivates M1?
-->

# Proposed models M1--M3

<!--
For each model:
  - State the model mathematically.
  - Justify every new term BEFORE showing results.
  - Show diagnostic plots and comment on improvement over the previous model.
-->

## Model M1: Trend and factor

Model M1, fitted for each day $i$ from 1 January 2020 to 31 December 2025, 
is specified as:

$$
y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \beta_3 t_i
+ \sum_{m=2}^{12} \gamma_m \mathbf{1}(m_i = m) 
+ \sum_{d=2}^{7} \delta_d \mathbf{1}(d_i = d) 
+ \varepsilon_i, \quad \varepsilon_i \sim N(0, \sigma^2)
$$

M1 extends M0 in three ways. First, month is switched from a numeric 
predictor to a factor, giving each month its own intercept. Figure 
\@ref(fig:fig-boxplotmonth) shows that the month--demand relationship 
is not monotonic, so forcing a single slope as M0 does is inappropriate. 
Second, day of week is added as a factor — the weekend binary in M0 
groups Saturday and Sunday together, but Figure 
\@ref(fig:fig-boxplotdow) shows these days behave differently, so a 
full factor allows each day its own effect. Third, a linear trend term 
$t_i$ is added to capture the drift in underlying demand visible in 
Figure \@ref(fig:fig-ts). Both factors use dummy encoding with January and Sunday as reference 
categories, so $\gamma_m$ and $\delta_d$ represent deviations from 
these baselines.



**Diagnostics of M1**

```{r fig-m1diagnostics,echo=FALSE,fig.cap="Diagnostics for M1",fig.width=7, fig.height=4}
m1_resid_fitted_plot + m1_qq_plot
```

Compared to M0, M1 shows some improvement — the fitted values span a wider range, suggesting the additional predictors (trend, factor month, factor day of week) give the model more flexibility to capture variation in demand. However, the core issues persist. The residuals versus fitted plot still shows increasing spread at higher fitted values, indicating that heteroscedasticity remains unresolved. The QQ plot continues to show a pronounced right tail, with standardized residuals reaching above 6 at the upper end. These issues motivate further extensions in M2 and M3. 

## Model M2: Nonlinear temperature

Model M2, fitted for each day $i$ from 1 January 2020 to 31 December 2025, 
is specified as:

$$
y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{1i}^2 + \beta_3 x_{2i} + \beta_4 t_i
+ \sum_{m=2}^{12} \gamma_m \mathbf{1}(m_i = m) 
+ \sum_{d=2}^{7} \delta_d \mathbf{1}(d_i = d) 
+ \varepsilon_i, \quad \varepsilon_i \sim N(0, \sigma^2)
$$

M2 extends M1 by replacing the linear temperature term with a quadratic, adding both $x_{1i}$ and $x_{1i}^2$ as predictors. Figure \@ref(fig:fig-scattertemp) shows that the temperature--demand relationship flattens above around 10°C, which a linear term cannot capture. The quadratic term allows the model to reflect this diminishing sensitivity, which is also physically plausible — once conditions are warm enough, further temperature increases have a smaller effect on cycling behaviour.

```{r fig-m2diagnostics,echo=FALSE,fig.cap="Diagnostics for M2",fig.width=7, fig.height=4}
m2_resid_fitted_plot+m2_qq_plot
```

M2 shows marginal improvement over M1 — residual spread is slightly more even and the lower QQ tail follows the reference line more closely. However, the right tail worsens, with standardised residuals exceeding 7, and heteroscedasticity persists. The consistently heavy right tail across M0, M1, and M2 suggests a Gaussian model is fundamentally inappropriate for the given data, motivating the log transformation in M3.


## Model M3: Log-Transformed Response

Model M3, fitted for each day $i$ from 1 January 2020 to 31 December 2025, 
is specified as:

$$
\log(y_i + 1) = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{1i}^2 + \beta_3 x_{2i} + \beta_4 t_i
+ \sum_{m=2}^{12} \gamma_m \mathbf{1}(m_i = m) 
+ \sum_{d=2}^{7} \delta_d \mathbf{1}(d_i = d) 
+ \varepsilon_i, \quad \varepsilon_i \sim N(0, \sigma^2)
$$

M3 keeps all predictors from M2 but log-transforms the response, fitting $\log(\text{count}_i + 1)$ rather than $\text{count}_i$ directly. The persistent right-skew and heteroscedasticity seen across M0--M2 are classic symptoms of modelling non-negative count data with a Gaussian assumption.The log transformation addresses this by compressing large values, reducing 
skewness and stabilising variance. The $+1$ ensures the transformation is defined if any zero counts are present.


```{r fig-m3diagnostics,echo=FALSE,fig.cap="Diagnostics for M3",fig.width=7, fig.height=4}
m3_resid_fitted_plot + m3_qq_plot
```

The diagnostic plots for M3 show a marked improvement — residuals are now scattered evenly around zero and the QQ plot follows the reference line closely across almost its entire range, compared to the severe right tail deviation in M2. Some deviation remains at both tails, indicating slight remaining non-normality, but the Gaussian assumption is considerably more plausible for M3 than any earlier model.



# Evaluation
## Statistical Performance


```{r tab-cv1, echo=FALSE}
kable(cv_table1,
      caption = "Leave-one-year-out cross-validation scores for models M0--M3. Lower is better for all scores.", booktabs = TRUE,
      linesep = "") %>%
  kable_styling(latex_options = "hold_position")
```

Table \@ref(tab:tab-cv1) shows the CV scores across all four models. RMSE and MAE improve consistently from M0 to M3, with M3 achieving the lowest values of 4531 and 3598 respectively. However, M0 achieves the lowest DS (18.0) and M3 the highest (18.5), with M3 also notably worse on IS. This disagreement arises because back-transforming M3's log-scale predictions inflates predictive variance via the delta method, penalising DS and IS. M3 is nonetheless selected as the final model — it achieves the best RMSE and MAE and is the only model whose diagnostic plots satisfy the normality and constant variance assumptions.


```{r tab-cv2, echo=FALSE}
kable(cv_table2,
      caption = "Monthly cross-validated RMSE and Dawid--Sebastiani score for the final model M3.", booktabs = TRUE,format = "latex") %>%
  kable_styling(latex_options = "hold_position")
```

Table \@ref(tab:tab-cv2) shows that predictive performance varies across the year. RMSE is lowest in July (3485.3) and highest in April (6485.7) and March (5886.6), suggesting early spring is hardest to predict. December has the worst DS (22.27), indicating poorer uncertainty calibration around that period. Early spring and December therefore present the greatest challenge for the final model.

<!--
Required content:
  1. Table 1: LOYO CV scores for M0-M3.
  2. Table 2: CV RMSE and DS by month for your final model.
  3. Discussion: do the scores agree on which model is best? 
     If not, explain why and state your final model choice explicitly.
-->

## Practical Performance

### Temperature effects and nonlinearity

```{r tab-tempcompare, echo=FALSE}
kable(temp_compare_table,
      caption = "Comparison of nonlinear temperature specifications in the M2 model family.", booktabs = TRUE,format = "latex") %>%
  kable_styling(latex_options = "hold_position")
```


Table \@ref(tab:tab-tempcompare) compares three nonlinear temperature specifications. `temp_min` performs best with the lowest RMSE (4574.7) 
and IS (25010.3), though DS is effectively identical across all three (difference of 0.03 at most). This is physically plausible — commuters are more likely to base cycling decisions on the coldest conditions of the day, making minimum temperature the most behaviourally relevant measure.

```{r tab-tempeffect, echo=FALSE}
kable(temp_effect_table,
      caption = "Estimated marginal effect of a 1°C increase using the chosen temperature variable.", booktabs = TRUE,format = "latex") %>%
  kable_styling(latex_options = "hold_position")
```  

Table \@ref(tab:tab-tempeffect) shows the marginal effects for the chosen specification. At 5°C, a 1°C increase is associated with approximately 53.9 additional cyclists per day. At 15°C, the model implies a decrease of 250.2 cyclists per day, though this more likely reflects the flattening curvature of the quadratic term at higher temperatures than a genuine reduction in cycling. As the model excludes a temperature $\times$ season interaction, this effect is assumed constant across the year.


### Seasonal prediction accuracy

<!--
One subsection per planning question (5.1-5.5 from the assignment brief).
Each must include mandatory fixed-format numbers and real-world interpretation.
-->


Table \@ref(tab:tab-cv2) shows that predictive performance varies 
noticeably across the year. RMSE is lowest in July (3485.3) and highest 
in April (6485.7) and March (5886.6), while December has the worst DS 
(22.27). Summer is easiest to predict as behaviour is stable and 
conditions consistently favourable. Early spring is hardest — as 
temperatures begin to improve, individuals differ in when they resume 
cycling, creating behavioural variation the model cannot capture. The 
poor performance in March, April, and December likely reflects both 
model limitations and missing predictors such as rainfall, wind, and 
public holidays.

### Long-term trend and extrapolation

```{r tab-trendcoefs, echo=FALSE}
kable(trend_coefs,
      caption = "Trend coefficients from models M1--M3.", booktabs = TRUE, format = "latex") %>%
  kable_styling(latex_options = "hold_position")
```

```{r tab-trendsig, echo=FALSE}
kable(trend_significance,
      caption = "Statistical significance of the trend coefficient in models M1--M3", booktabs = TRUE, format = "latex") %>%
  kable_styling(latex_options = "hold_position")
```

```{r tab-growthrate, echo=FALSE}
kable(growth_rate_table,
      caption = "Annual growth rate expressed relative to the predicted count on 1 January 2020.", booktabs = TRUE, format = "latex") %>%
  kable_styling(latex_options = "hold_position")
```

Tables \@ref(tab:tab-trendcoefs) and \@ref(tab:tab-trendsig) show that all three models estimate a statistically significant negative trend. 
M1 implies 692 fewer cyclists per year, M2 683, and M3 786, corresponding to annual percentage changes of -5.66%, -5.48%, and -6.82% relative to the predicted count on 1 January 2020. Table \@ref(tab:tab-growthrate) shows the trend is consistent across models, though the slightly more 
negative estimate in M3 suggests skewness and non-constant variance shift the implied rate somewhat.

```{r tab-trendcomp, echo=FALSE}
kable(trend_comparison,
      caption = "Comparison of M3 trend estimates with and without a COVID indicator.", booktabs = TRUE, format = "latex") %>%
  kable_styling(latex_options = "hold_position")
```

Table \@ref(tab:tab-trendcomp) shows that including a COVID indicator shifts the M3 trend from -2.152 to -2.904 cyclists/day (-786 to -1060 
per year), so the trend should not be read as evidence of structural decline.

```{r fig-extrapolation, echo=FALSE,fig.width=7, fig.height=2.5, fig.cap="Extrapolation of Cycling Demand"}
extrapolation_plot
```

```{r tab-threshold, echo=FALSE}
kable(threshold_15000,
      caption = "Earliest projected date at which predicted summer weekday demand exceeds 15,000 cyclists/day.", booktabs = TRUE,format = "latex") %>%
  kable_styling(latex_options = "hold_position")
```
The extrapolation based on M3 projects decline in Figure \@ref(fig:fig-extrapolation), rather than an increase, summer weekday cycling demand through to 2035. Using the fitted model we can see that the predicted demand is already above the 15,000 cyclists/day threshold at the start of the period, starting at an estimated value of 15,579 on 1 January 2020 as seen in Table \@ref(tab:tab-threshold). It then falls steadily, where in 15 years, the estimated value for cyclists/day is 5,300. This means that, at the current fitted rate, the model does not predict future growth toward the threshold; instead, it suggests that demand moves away from it over time.

This kind of long-run extrapolation should be treated with caution for at least two reasons. First, the fitted linear trend is estimated over a short period, specifically the COVID-19 disruption, where many were forced to stay indoors, hence it may not represent the true long-term path of cycling demand. Secondly, the projection holds weather and seasonal conditions fixed and ignores other factors that could substantially alter future demand. Notably as infrastructure investment, policy changes, fuel prices, or changes in population and commuting patterns. For these reasons, the extrapolation is useful as a simple model-based scenario, but not as a reliable long-term forecast.

### Identifying a poorly-fit period

This subsection examines the residuals of the final model over time in order to identify a specific period in which the model performs notably poorly. The aim is to determine whether the model systematically over- or under-predicts during that period, and to consider what that implies from both a statistical and transport-planning perspective.

```{r fig-restime, echo=FALSE,fig.width=7, fig.height=4, fig.cap="Residuals over Time"}
resid_time_plot
```

Figure \@ref(fig:fig-restime) identifies periods where the final model appears to fit particularly poorly. This can be caused either by residuals that are unusually large in magnitude or because they are persistently positive or negative over a sustained interval.

```{r tab-poorperiod, echo=FALSE}
kable(head(poor_period_table, 10),
      caption = "Months with the largest RMSE under the final model.", booktabs = TRUE,format = "latex") %>%
  kable_styling(latex_options = "hold_position")

```

Table 8 shows the months in which the final model performs worst, ranked by period-specific RMSE. This table will therefore help indicate the exact month with worst performance.

From Table \@ref(tab:tab-poorperiod), we find that April 2022 stands out as the single poorest-fitted month. It has the largest month-specific RMSE in the table, equal to 11317.8, and a positive mean residual of 8934.4. This indicates that the model under-predicts cyclist demand consistently during this period of time rather than making a few isolated large errors.

This implies April 2022 was a period of unusually high cycling demand that is not fully explained by the variables included in the model. The residual plot shows a large grouping of positive residuals during spring and summer 2022. This suggests that the model may be missing some factor that lifted counts above their usual seasonal level. 


From a transport-planning perspective, this reflect changes in travel behavior, especially if cycling demand temporarily increased because of favorable or unfavorable conditions. There was potentially a post-pandemic behavioural shifts, where increased free movement was taken advantage. On the other hand, there may have been external factors such as fuel prices. The Ukraine-Russia war started in Feb.2022 and had resulted in significant increases in the prices of crude oil. Statistically, the poor fit also reflects model limitations: the final model includes temperature, seasonality, day of week, and trend, but omits potentially important short-term drivers such as rainfall, wind, holidays, oil prices and special events. April 2022 therefore appears to be a period where both the data limitations and the model specification contribute to poor predictive performance.

### Original Diagnostic Plot

This subsection presents an original plot that is not a standard diagnostic, designed to reveal an interesting feature of the final model that is not visible from standard diagnostics plots. This plot tracks the monthly mean residual over time, showing whether the model systematically over- or under-predicts demand in different periods.

```{r fig-predictionbias, echo=FALSE,fig.width=7, fig.height=2.5, fig.cap="Monthly Mean Prediction Bias"}
original_bias_plot
```

The plot in Figure \@ref(fig:fig-predictionbias) shows the monthly mean residual of the final model over time. Positive values  indicate systematic under-prediction and negative values indicate systematic over-prediction. Multiple periods are visibly  out, particularly parts of 2022 and late 2024, where the model persistently under-predicts cycling demand, while much of 2025 shows negative mean residuals, indicating over-prediction. A transport planner should conclude that the model captures overall demand patterns reasonably well, but it struggles to track shifts in underlying behaviour across time. In particular, some periods appear to involve sustained changes in demand that are not fully explained by temperature, seasonality, day of week, and trend alone.

This plot is useful because it highlights regional bias rather than just overall error magnitude. Unlike RMSE or QQ plots, it shows when the model is persistently wrong in one direction, which is especially important for planning. A model that repeatedly under-predicts demand during certain periods may lead planners to underestimate pressure on cycling infrastructure, while persistent over-prediction could lead to over-allocation of resources.

# Discussion and limitations


<!--
3-5 paragraphs covering:
  - Whether model assumptions appear satisfied.
  - Key limitations of the model and data.
  - What ideas would improve prediction.
  - Whether findings generalise to other cities or future years.
Do not introduce new results here.
-->

Of the four models, M3 best satisfies linear regression assumptions 
though none fully does — M0 through M2 all show indications of 
heteroscedasticity and a heavy right tail, and while the log 
transformation in M3 largely addresses both, some tail deviation remains.

The dataset's main limitation is that it only covers temperature and 
calendar variables, omitting rainfall and wind which are known to 
influence cycling decisions. The response variable also aggregates counts 
across all stations, preventing the model from distinguishing commuter 
and leisure routes.

The negative trend should be interpreted cautiously. A single linear 
trend across 2020--2025 conflates the COVID dip, recovery, and later 
years, and as shown in Section 5.3, the estimate shifts materially 
depending on how the pandemic is treated.

The most practical improvements would be adding precipitation and wind 
speed, replacing the linear trend with a piecewise specification, and 
using a Poisson or negative binomial model to avoid the back-transformation 
issue affecting M3's DS and IS scores. Disaggregating to individual 
stations would also allow more targeted route-level conclusions.

These findings should not be generalised beyond Edinburgh without caution 
— the temperature effects, commuter patterns, and trend reflect a specific 
city over an unusual period. The 2035 extrapolation in particular should 
be treated as an illustrative scenario rather than a forecast.
# Appendix: Code


All analysis code is in `code.R` and is reproduced below.

```{r print-code, echo=TRUE, eval=FALSE, code=readLines("Proj2code.R")}
```


```{r}

```

